\parindent = 0pt
\timesrm
\c{\bigrmb References}
\vskip 15pt
Allan, R.S., Charles, G.E. \& Mason, S.G. 1961 The approach of gas bubbles to a 
gas/liquid interface. {\sl J. Coll. Sci.}, {\bf 16} pp~150--165.
\vskip 8pt
Al-Rubeai, M. Chalder, S. \& Emery, A.N. 1990 Dynamic assay of synthetic
activities in single hybridoma cells. In {\sl Production of 
Biologicals from Animal Cells in Culture.} (Eds. Spier, R.E., 
Griffith, J.B. \& Meigner, B.) pp~587--593. Butterworth Heinemann.
\vskip 8pt 
Atkinson, K. Graham, I. \& Sloan, I. 1983 Piecewise continuous collocation
for integral equations. {\sl SIAM J. Numer. Anal.} {\bf 20(1)} pp~172--186.
\vskip 8pt
Baker, G.R., Meiron, D.I. \& Orszag, S.A. 1982 Generalised vortex methods
for free-surface flow problems. {\sl J. Fluid Mech.}, {\bf 123} pp~477--501.
\vskip 8pt
Baker, G.R. \& Moore, D.W. 1989 The rise and distortion of a two-dimensional
bubble in an inviscid liquid. {\sl Phys. Fluids A}, {\bf 1(9)} pp~1451--1459.
\vskip 8pt
Batchelor, G.K. 1987 The stability of a large gas bubble rising 
through liquid. {\sl J. Fluid Mech.}, {\bf 184} pp~399--422.
\vskip 8pt
Batchelor, G.K. 1967 {\sl An Introduction to Fluid Dynamics.} Cambridge
University Press.
\vskip 8pt
Baumel, R.T., Burley, S.K., Freeman, D.F., Gammel, J.K. \& Nuttall, J. 1982
The rise of a cylindrical bubble in an inviscid liquid. {\sl Can J. Phys.},
{\bf 60} pp~999--1007.
\vskip 8pt
Bavarian, F., Fan, L.S. \& Chalmers, J.J. 1991 Microscopic visualization
of insect cell-bubble interactions I: Rising bubbles, air-medium interface, 
and the foam layer. {\sl Biotech. Prog.}, {\bf 7(2)} pp~140--150.
\vskip 8pt
Best, J.P. \& Kucera, A. 1992 A numerical investigation of
nonspherical rebounding bubbles. {\sl J. Fluid Mech.}, {\bf 245} pp~137--154.
\vskip 8pt
Bikerman, J.J. 1973 {\sl Foams.} Springer-Verlag.
\vskip 8pt
Bhaga, D. \& Weber, M.E. 1981 Bubbles in viscous liquids:
shapes, wakes and velocities. {\sl J. Fluid Mech.}, {\bf 105} pp~61-85.
\vskip 8pt
Blake, J.R. 1988 The Kelvin impulse: application to cavitation 
bubble dynamics. {\sl J. Austral. Math. Soc. B}, {\bf 30} pp~127-146.
\vskip 8pt
Blake, J.R. \& Gibson, D.C. 1981 Growth and collapse of a vapour 
cavity near a free surface. {\sl J. Fluid Mech.}, {\bf 111} 
pp~123--140.
\vskip 8pt
Blake, J.R. \& Kucera, A. 1988 Coning in oil reservoirs {\sl Math. Sci.}, {\bf 
13} pp~36--47.
\vskip 8pt
Blake, J.R., Taib, B.B. \& Doherty, G. 1986 Transient cavities near boundaries.
Part I. Rigid Boundary. {\sl J. Fluid Mech.}, {\bf 170} pp~479--497.
\vskip 8pt
Blake, J.R., Taib, B.B. \& Doherty, G. 1987 Transient cavities near boundaries.
Part II. Free Surface. {\sl J. Fluid Mech.}, {\bf 181} pp~197--212.
\vskip 8pt
Blanchard, D.C. \& Syzdek, L.D. 1972 Concentration of bacteria in jet drops 
from bursting bubbles. {\sl J. Geophys. Res.}, {\bf 77(27)} pp~5087--5099.
\vskip 8pt
Boris, J.P. 1989 New directions in computational fluid dynamics. 
{\sl Ann. Rev. Fluid
Mech.}, {\bf 21} pp~345--38
\vskip 8pt
Brebbia, C.A. \& Walker, S.J. 1980 {\sl Boundary Element techniques in
Engineering.} Butterworths.
\vskip 8pt
Chahine, G.L. 1991 Dynamics of the interaction of non-spherical
cavities. In {\sl Mathematical approaches in hydrodynamics}, 
(Ed. Miloh, T.), SIAM Philadelphia, pp~51--67.
\vskip 8pt
Chalmers, J.J. 1992 Private communication.
\vskip 8pt
Chalmers, J.J. \& Bavarian, F. 1991
Microscopic visualization of insect cell-bubble interactions II: The film and 
bubble
rupture. {\sl Biotech. Prog.}, {\bf 7(2)} pp~151--159.
\vskip 8pt
Clift, T, R., Grace, J.R. \& Weber, M.E. 1978 {\sl Bubbles, Drops, and Particles.}
Academic Press.
\vskip 8pt
Culick, F.E.C. 1960 Comments on a ruptured soap film. {\sl J. Appl. Phys.},
{\bf 31I} pp~1128--1129.
\vskip 8pt
Fink, P.T. \& Soh, W.K. 1978 A new approach to roll-up 
calculations of vortex sheets. {\sl Proc. Roy. Soc. Lond. A}, 
{\bf 362} pp~195--209.
\vskip 8pt
Forbes, L.K. \& Hocking, G.C. 1990 Flow caused by a point sink in 
a fluid having a free surface. {\sl J. Austral. Math. Soc. B}, 
{\bf 32} pp~231--249.
\vskip 8pt
Garner, F.H., Ellis, S.R.M. \& Lacey, J.A. 1954 The size distribution and
entrainment of droplets. {\sl Trans. Instn. Chem. Engrs.}, {\bf 32} 
pp~222--235.
\vskip 8pt
Guerri, L., Lucca, G. \& Prosperetti, A. 1981 A numerical method for
the dynamics of non-spherical cavitation bubbles. {\sl Proc. 2nd
Int. Colloq. on drops and bubbles. California.} pp~175--181.
\vskip 8pt
Hahn P.-S., Chen, J.-D. \& Slattery, J.C. 1985 Effects of London-van 
der Waals forces on the thinning and rupture of a dimpled liquid
film as a small drop or bubble approaches a fluid-fluid 
interface. {\sl AIChE. J.}, {\bf 31(12)} pp~2026--2038.
\vskip 8pt
Handa, A. 1986 {\sl Gas-Liquid Interfacial Effects on the Growth of
Hybridomas and Other Suspended Mammalian Cells.} Ph.D. Thesis. 
University of Birmingham, UK.
\vskip 8pt
Handa, A., Emery, A.N. \& Spier, R.E. 1987 On the evaluation of gas-liquid
interfacial effects on hybridoma viability in bubble column bioreactors.
{\sl Dev. Biol. Standard.}, {\bf 66} pp~241--253.
\vskip 8pt
Harper, J.F. 1972 The motion of bubbles and drops through liquids.
{\sl Adv. Appl. Mech.}, {\bf 12} pp~59-129.
\vskip 8pt
Harris, P. 1992 A numerical model for determining the motion of a 
bubble close to a fixed rigid structure in a fluid.
{\sl Int. J. Numer. Meth. Engng.}, {\bf 33(9)} pp~1813-1822.
\vskip 8pt
Hastings, C. 1955 {\sl Approximations For Digital Computers.} Princeton 
University Press, Princeton, N.J.
\vskip 8pt
Havelock, T.H. 1927 The method of images in some problems of 
surface waves. {\sl Proc. Roy. Soc. Lond. A}, {\bf 115} pp~268--280.
\vskip 8pt
Ivanov, B.I., Kralchevsky, P.A. \& Nikolov, A.D. 1986 Film 
and line tension effects
on the attachment of particles to an interface. Parts I--IV. 
{\sl J. Coll. Int. Sci.}, {\bf 102(1)} pp~97--143.
\vskip 8pt
Jaswon, M.A. \& Symm, G.T. 1963 Integral equation methods in potential theory.
Parts I and II. {\sl Proc. Roy. Soc. Lond. A}, {\bf 275} pp~23--46.
\vskip 8pt
Jaswon, M.A. \& Symm, G.T. 1977 {\sl Integral Equation Methods in Potential
Theory and Elastostatics.} Academic Press.
\vskip 8pt
Kientzler, C.F., Arons, A.B., Blanchard, D.C. \& Woodcock, A.H. 1954
Photographic investigation of the projection of droplets by bubbles 
bursting at a water surface. {\sl Tellus}, {\bf 6(1)} pp~1--7.
\vskip 8pt
King, A.C. \& Bloor, M.I.G. 1988 A note on the free surface 
induced by a submerged source at infinite Froude number. {\sl J. 
Austral. Math. Soc. B}, {\bf 30} pp~147--156.
\vskip 8pt
Kioukia, N. 1990. {\sl Physical and Chemical Factors Affecting Hybridoma
Growth.} M.Sc. Dissertation. University of Birmingham, UK.
\vskip 8pt
Kirkpatrick, R.D. \& Lockett, M.J. 1974 The influence of 
approach velocity on bubble
coalescence. {\sl Chem. Eng. Sci.}, {\bf 29} pp~2363--2373.
\vskip 8pt
Kowalski, A.J. 1991 {\sl Microscale fluid dynamic effects in suspension processing
and attrition of cell cultures.} Ph.D. Thesis. University of Birmingham, UK.
\vskip 8pt 
Kre$\beta$, R. \& Spassov, W.T. 1983 On the condition number of
boundary integral operators for the exterior Dirichlet problem
for the Helmholtz equation. {\sl Numer. Math.}, {\bf 42} pp~77--95.
\vskip 8pt
Kucera, A. 1991 A boundary integral method applied to the growth and collapse
of bubbles near a rigid boundary. To appear in {\sl J. Comp. Phys.}
\vskip 8pt
Kunas, K.T. \& Papoutsakis, E.T. 1989 Increasing serum concentrations
decrease cell death and allow cell growth of hybridoma cells at higher
agitation rates. {\sl Biotech. Letts.}, {\bf 11(8)} pp~525--530.
\vskip 8pt
Kunas, K.T. \& Papoutsakis, E.T. 1990 Damage mechanisms of suspended animal
cells in agitated bioreactors with and without bubble entrainment.
{\sl Biotech. and Bioeng.}, {\bf 36} pp~476--483.
\vskip 8pt
Lamb, H. 1932 {\sl Hydrodynamics.} Dover.
\vskip 8pt
Levich, V.G. 1949 Motion of gaseous bubbles with high Reynolds numbers.
{\sl Zhn. Eksp. Teor. Fiz.}, {\bf 19} pp~18--24.
\vskip 8pt
Levich, V.G. 1962 {\sl Physicochemical Hydrodynamics.} Prentice-Hall, 
New Jersey.
\vskip 8pt
Lighthill, M.J. 1986 {\sl An Informal Introduction to Fluid Mechanics.}
Oxford.
\vskip 8pt
Lo, L.L. 1983 The meniscus on a needle --- a lesson in matching. 
{\sl J. Fluid Mech.},
{\bf 132} pp~85--78.
\vskip 8pt
Longuet-Higgins, M.S. 1980 On the forming of sharp corners at a free surface.
{\sl Proc. Roy. Soc. Lond. A}, {\bf 371} pp~453--478.
\vskip 8pt
Longuet-Higgins, M.S. 1983 Bubbles, breaking waves and hyperbolic jets 
at a free surface. {\sl J. Fluid Mech.}, {\bf 127} pp~103--121.
\vskip 8pt
Longuet-Higgins, M.S. \& Cokelet, E.D. 1976 The deformation
of steep surface waves on water. I. A numerical method of computation.
{\sl Proc Roy. Soc. Lond. A}, {\bf 350} pp~1--26.
\vskip 8pt
Lundgren, T.S. \& Mansour, N.N. 1988 Oscillations of drops in zero gravity with
weak viscous effects. {\sl J. Fluid Mech.}, {\bf 194} pp~479--510.
\vskip 8pt
Lundgren, T.S. \& Mansour, N.N. 1991 Vortex ring bubbles. {\sl J. Fluid Mech.},
{\bf 224} pp~177--196.
\vskip 8pt
MacIntyre, F. 1972 Flow patterns in breaking bubbles. {\sl J. 
Geophys. Res.}, {\bf 77(27)} pp~5211--5228.
\vskip 8pt
Martin, P.A. \& Rizzo, F.J. 1989 On boundary integral equations for crack
problems. {\sl Proc. Roy. Soc. Lond. A}, {\bf 421} pp~341--355.
\vskip 8pt
McQueen, A. \& Bailey, J.E. 1989 Influence of serum level, cell line, flow type
and viscosity on flow-induced lysis of suspended mammalian cells. {\sl 
Biotech. Letts.}, {\bf 11(8)} pp~531--536.
\vskip 8pt
Mikhlin, S.G. 1957 {\sl Integral Equations.} Pergamon.
\vskip 8pt
Miksis, M., Vanden Broeck, J.-M. \& Keller, J.B. 1981 Axisymmetric 
bubble or drop in a uniform flow. {\sl J. Fluid Mech.}, {\bf 108} 
pp~89--100.
\vskip 8pt
Miksis, M., Vanden Broeck, J.-M. \& Keller, J.B. 1982 Rising 
bubbles. {\sl J. Fluid Mech.}, {\bf 123} pp~31--41.
\vskip 8pt
Moore, D.W. 1963 The boundary layer on a spherical gas bubble.
{\sl J. Fluid Mech.}, {\bf 16} pp~161--176.
\vskip 8pt
Moore, D.W. 1965 The velocity of rise of distorted gas bubbles in a 
liquid of small viscosity. {\sl J. Fluid Mech.}, {\bf 23(4)} pp~749--766.
\vskip 8pt
Moore, D.W. 1973 A numerical study of the roll-up of a finite 
vortex sheet. {\sl J. Fluid Mech.}, {\bf 63(2)} pp~225--235.
\vskip 8pt
Moore, D.W. 1981 On the point vortex method. {\sl J. Sci. Stat. 
Comp.}, {\bf 2(1)} pp~65--84.
\vskip 8pt
Muskhelishvili, N.I. 1953 {\sl Singular Integral Equations.} Noordhoff.
\vskip 8pt
Newitt, D.M., Dombrowski, N. \& Knelman, F.H. 1954 Liquid entrainment 1. The 
mechanism
of drop formation from gas or vapour bubbles. {\sl Trans. Instn. Chem.
Engrs.}, {\bf 32} pp~244--261.
\vskip 8pt
Oguz H.N. \& Prosperetti, A. 1989 Surface tension effects in the 
contact of liquid surfaces. {\sl J. Fluid Mech.}, {\bf 203} pp~149--171.
\vskip 8pt
Oguz H.N. \& Prosperetti, A. 1990a A generalization of the impulse and virial
theorems with an application to bubble oscillations. 
{\sl J. Fluid Mech.}, {\bf 218} pp~143-162.
\vskip 8pt
Oguz H.N. \& Prosperetti, A. 1990b Bubble entrainment by the 
impact of drops on liquid surfaces. {\sl J. Fluid Mech.}, {\bf 
219} pp~143--179. 
\vskip 8pt
Oh, S.K.W., Nienow, A.W., Al-Rubeai, M. \& Emery, A.N. 1989 The effects of
agitation intensity with and without continuous sparging on the growth 
and antibody production of hybridoma cells. {\sl J. Biotech.}, {\bf 12}
pp~45--62.
\vskip 8pt
Oh, S.K.W., Nienow, A.W., Al-Rubeai, M. \& Emery, A.N. 1992 
Further studies of the culture
of mouse hybridomas in an agitated bioreactor with and
without continuous sparging.
{\sl J. Biotech.}, {\bf 22} pp~245--270. 
\vskip 8pt
Orton, D.R. \& Wang, D.I.C. 1990 Effects of gas interfaces
on animal cells in bubble aerated bioreactors. Paper No. 103B 
presented at AIChE Annual Meeting, Chicago.
\vskip 8pt
Princen, H.M. 1963 Shape of a fluid drop at a liquid-liquid interface.
{\sl J. Coll. Sci.}, {\bf 18} pp~178--195.
\vskip 8pt
Ranz, W.E. 1959 Some experiments on the dynamics of liquid films. {\sl J. Appl. 
Phys.}, {\bf 30(12)} pp~1950-1955.
\vskip 8pt
Rayleigh, Lord 1878 On the instability of jets. {\sl Proc. Lond. Math. Soc.},
{\bf 10} pp~4-13.
\vskip 8pt
Rayleigh, Lord 1891 Some applications of photography. {\sl Nature.}, 
{\bf 44} pp~249--254.
\vskip 8pt
Resch, F.J., Darrozes, J.S. \& A\`feti, G.M. 1986 Marine liquid aerosol production
from bursting of air bubbles. {\sl J. Geophys. Res.}, {\bf 91(C1)} pp~1019-1029.
\vskip 8pt
Robinson, P.B. 1992 {\sl The numerical modelling of bubble interactions.}
M.Sc. Thesis, University of Birmingham, U.K.
\vskip 8pt
Robinson, P.B., Boulton-Stone, J.M. \& Blake, J.R. 1993
The interaction of 
two-dimensional gas bubbles. Submitted to {\sl J. Eng. Math.}
\vskip 8pt
Ryskin, G. \& Leal, L.G. 1984 Numerical solution of free-boundary problems
in fluid mechanics. Part 2. Buoyancy-driven motion of a gas bubble through
a quiescent liquid. {\sl J. Fluid Mech.}, {\bf 148} pp~19--35.
\vskip 8pt
Scheludko, A. 1962 Certain particulars of foam layers. II
Kinetic stability, critical thickness and equilibrium thickness.
{\sl Proc. K. Ned. Akad. Wetensch B}, {\bf 65(1)} 87--96.
\vskip 8pt
Soh, W.K. 1987 Boundary element technique for flow vizualization studies of
nonlinear water waves. {\sl Eng. Anal.}, {\bf 4(2)} pp~106--112.
\vskip 8pt
Stoker, J.J. 1957 {\sl Water Waves. The Mathematical Theory with Applications.} 
Interscience Publishers.
\vskip 8pt
Stoos, J.A. \& Leal, L.G. 1990 A spherical particle straddling a fluid/gas
interface in an axisymmetric straining flow. {\sl J. Fluid Mech.}, 
{\bf 217} pp~263-298.
\vskip 8pt
Taib, B.B. 1985 {\sl Boundary Element Method Applied to Cavitation Bubble
Dynamics.} Ph.D. Thesis. University of Wollongong, Australia.
\vskip 8pt
Taylor, G.I. \& Michael, D.H. 1973 On making holes in a sheet of fluid. 
{\sl J. Fluid Mech.}, {\bf 58(4)} pp~625--639.
\vskip 8pt
Taylor, T.D. \& Acrivos, A. 1964 On the deformation and drag of a falling
viscous drop at low Reynolds number. {\sl J. Fluid. Mech.}, {\bf 18} 
pp~466-476.
\vskip 8pt
Tramper, J., Williams, J.B. \& Joustra, D. 1986 Shear sensitivity of 
insect cells in suspension. {\sl Enzyme Microb. Technol.}, {\bf 8} pp~33--36.
\vskip 8pt
Tryggvason, G. 1988 Deformation of a free surface as a result of 
vortical flows. {\sl Phys. Fluids}, {\bf 31(5)} pp~955--957.
\vskip 8pt
Tuck, E.O. \& Vanden Broeck, J.-M. 1984 A cusp-like free-surface 
flow due to a submerged source or sink. {\sl J. Austral. Math. 
Soc. B}, {\bf 25} pp~443--450.
\vskip 8pt
Unverdi, S.O. \& Tryggvason, G 1992 A front-tracking method for viscous, 
incompressible, multi-fluid flows. {J. Comp. Phys.}, {\bf 100(1)} pp~25--37.
\vskip 8pt
Vanden Broeck, J.-M., Schwartz, L.W. \& Tuck, E.O. 1978 Divergent 
low-Froude-number series expansions of nonlinear free-surface 
flow problems. {\sl Proc. Roy. Soc. Lond. A}, {\bf 361} pp~
207--224.
\vskip 8pt
Vrij, A. 1966 Possible mechanism for the spontaneous rupture of 
thin, free liquid films. {\sl Disc. Faraday Soc.}, {\bf 42} pp~
23--33.
\vskip 8pt
Walters, J.K. \& Davidson, J.F. 1962 The initial motion of a gas bubble
in an inviscid liquid. Part I. The two-dimensional bubble. {\sl J. Fluid
Mech.}, {\bf 12} pp~408--417.
\vskip 8pt
Wehausen, J.V \& Laitone, E.V. 1960 Surface waves. {\sl Handbuch der 
Physik}, {\bf 9(3)} pp~446--778.
\vskip 8pt
Yang, J.-D., Wang, N.S., Chang, K.-C. \& Calabrese, R.V. 1990
Hybridoma cell inactivation by air sparging in a mechanically agitated
bioreactor. Submitted for publication.
\vskip 8pt
Zhang, Z., Ferenczi, M.A., Lush, A.C. \& 
Thomas, C.R. 1991 A novel 
micromanipulation technique for measuring the bursting strength of
single mammalian cells. {\sl Appl. Microbiol. Biotechnol.},
{\bf 36} pp~208--210.
